Getting the Binomial RV from the Bernoulli
========================================================
In class we defined the Binomial$(n,p)$ random variable as the sum of $n$ independent Bernoulli$(p)$ random variables. In other words, the Binomial$(n,p)$ equals the total number of successes (ones) in $n$ independent Bernoulli trials, each with probability of success (one) equal to $p$. The point of this document is to convince you that this definition actually makes sense and really does lead to the formulas from class.
***
A complete transcript of the R source code for this document is available on my [Github Page](https://gist.github.com/fditraglia/7011503)
***
## A Function to Simulate Bernoulli Random Variables
R doesn't have a built-in function to simulate Bernoulli Random Variables since it treats them as a Binomial$(n=1,p)$ random variable. Let's make our own:
```{r}
#This function simulates n independent, identically distributed Bernoulli Random Variables
rbern <- function(n, p){
  
  sims <- sample(0:1, size = n, replace = TRUE, prob = c(1-p, p))
  return(sims)
  
}
```
The argument ``prob = c(1-p, p)`` is new. It tells ``sample`` to draw ``0`` with probability ``1-p`` and ``1`` with probability ``p``. This is exactly what we want. Let's test it out:
```{r}
rbern(30, 0.1)
rbern(30, 0.5)
rbern(30, 0.9)
```
## A Function to Simulate the Sum of Bernoullis
The function ``rbern`` returns all the *individual* Bernoulli draws. What we need to make a connection with the Binomial RV is a function that returns the *sum* of these draws instead. This is easy: we just call ``rbern`` and take its sum:
```{r}
#This function simulates n independent, identically distributed Bernoulli Random Variables and returns their sum
rbern.sum <- function(n, p){
  
  sims <- rbern(n, p)
  return(sum(sims))
  
}
```
Let's test it out:
```{r}
rbern.sum(50, 0.1)
rbern.sum(50, 0.5)
rbern.sum(50, 0.9)
```
## Finally the Binomial RV
It turns out that our function ``rbern.sum`` makes a *single* random draw from a Binomial$(n,p)$ random variable. How do I know this? We constructed it exactly following the definition of the Binomial from class: draw the some Bernoullis and sum them up. But don't take my word for it. Let's verify this with a simulation.

### Simulating 100,000 Binomial$(n = 10, p = 0.5)$ Draws
To do this, we'll use ``rbern.sum`` and ``replicate``
```{r}
binom.sims <- replicate(10^5, rbern.sum(10, 0.5))
head(binom.sims)
```
Let's make a plot of the relative frequencies from this simulation experiment:
```{r}
p.sim <- table(binom.sims)/10^5
p.sim
plot(p.sim, type = 'h')
```
What we've done here is approximate a pmf using Monte Carlo simulation. Now let's compare it to the *actual* pmf of a Binomial RV.

### The Exact pmf for a Bernoulli$(n =10, p=0.5)$ 
We know how to calculate this using what we covered in R Tutorial #5
```{r}
support <- 0:10
p.true <- dbinom(support, size = 10, prob = 0.5)
rbind(support, p.true)
plot(support, p.true, type = 'h')
```
The plot of the *actual* pmf and the associated probabilities match those from the simulation *almost perfectly*:
```{r}
p.sim - p.true
```
As you can see, the differences are *tiny*. They'd be even smaller if we used more replications.
***
**I hope this example has given you a bit more intuition about the Binomial RV. I'm not going to assign any exercises here, but it would be a good idea to try some simulation experiments of your own with different values for $n$ and $p$.**
***




