---
title: "R Tutorial #2 -- Econ 103"
output:
  html_document:
    toc: true
    theme: readable
    css: custom.css
  pdf_document:
    toc: true
    highlight: zenburn
---
This tutorial has two parts. In the first part, you'll learn how to implement the summary statistics and graphics from class using R. In the second part, you'll use what you've learned to explore a real-world dataset: the passenger manifest from the Titanic.


## Summary Stats and Graphics

### Load the Data
In this section we'll work with data from the student survey. Since results for this semester haven't been collected yet, we'll start with data from a previous semester:
```{r}
survey <- read.csv("http://www.ditraglia.com/econ103/old_survey.csv")
```
For this example we'll only need the first six columns of the survey so I'll first drop the others:
```{r}
survey <- survey[,1:6]
```

It's always a good idea to take a look at your data after loading it. The functions ``head`` and ``tail``, introduced in the previous tutorial, make this easy:
```{r}
head(survey)
tail(survey)
```

### ``hist`` 
The command for a histogram in R is ``hist``. Its input has to be a vector, not a dataframe. This means you'll need to select the column you want using what you learned in R Tutorial #1. For example, we can make a histogram of the column handedness as follows:
```{r}
hist(survey$handedness)
```

The default title and axis label aren't very nice. We can change them using the following options
```{r}
hist(survey$handedness, xlab = 'Handedness Score', main = 'Histogram of Handedness Scores', ylab = '# of Students')
```

Note that R doesn't care if you use single or double quotes as long are you're consistent (the symbol you start with is the symbol you end with). 

You can change the number of bins via the argument ``breaks``
```{r}
hist(survey$handedness, breaks = 20, xlab = 'Handedness Score', main = 'Histogram of Handedness Scores')
```

By default, R produces histograms in terms of *frequencies*. That is, it counts the *numbers* of observations in each bin. To change this to *relative frequencies*, that is proportions, we set the argument ``freq`` to ``FALSE`` 
```{r}
hist(survey$handedness, breaks = 20, freq = FALSE, xlab = 'Handedness Score', main = 'Histogram of Handedness')
```

Notice that if you don't supply a label for the y-axis, R defaults to "Frequency" unless you set ``freq = FALSE``, in which case it uses "Density." We'll learn about densities later in the course.

### ``plot``
The general-purpose command for plotting in R is rather unimaginatively called ``plot``. This is an incredibly powerful command, but we'll mostly use it to make simple scatter plots. To plot height on the the x-axis and handspan on the y-axis, we use the following command:
```{r}
plot(survey$height, survey$handspan)
```

Again, these arguments need to be vectors so we select the desired columns of our dataframe. If you change the order, you reverse the plot:
```{r}
plot(survey$handspan, survey$height)
```

You can use the same arguments for ``plot`` as you did for ``hist`` to change the title and axis labels and make the plot look cleaner:
```{r}
plot(survey$height, survey$handspan, xlab = "height (in)", ylab = "handspan (cm)")
```

Plot has many different arguments that you can use to customize the appearance of your plot. We can change the color of the points:
```{r}
plot(survey$height, survey$handspan, xlab = "height (in)", ylab = "handspan (cm)", col = "red")
```

the shape of the points:
```{r}
plot(survey$height, survey$handspan, xlab = "height (in)", ylab = "handspan (cm)", col = "red", pch = 3)
```

and we can even plot connected line segments rather than points, although this isn't really useful for the current example:
```{r}
plot(survey$height, survey$handspan, xlab = "height (in)", ylab = "handspan (cm)", col = "red", pch = 3, type = 'l')
```

Have a look at the help file for plot and try out some of the other options.


### ``pairs``
The function ``pairs`` is a great way to visualize many columns of a dataframe at once. Unlike ``plot``, whose two first arguments need to be vectors, ``pairs`` can take a *whole dataframe* as its argument. It produces a kind of scatter plot on steroids, in which every column in the dataframe is plotted against *every other column*. Let's try it out with some of the measurements from the survey:
```{r}
survey.measurements <- survey[, c("handedness", "handspan", "height")]
pairs(survey.measurements)
```

There seems to be a strong positive relationship between handspan and height but not much of a relationship between handeness and either of the other columns.

It takes a bit of practice to be able to read a pairs plot. Spend a little time looking at this one to figure it out and look at the help file if needed. 

### ``boxplot``
A boxplot is an alternative way to visualize the distribution of a dataset. The command in R is just what you'd expect, and you can use the same labels as above. The first argument must be a vector:
```{r}
boxplot(survey$handspan, ylab = "Handspan(cm)")
```

The "box" in a boxplot shows the middle 50% of the data. The thick line in the middle is the median, and the lines immediately above and below are the 25th and 75th percentiles. This means that the width of the box equals the interquartile range of the data.

Traditionally, the "whiskers" in a boxplot show the maximum and minimum of the data, but R has an interesting special feature. If there are any observations that are extremely unusual (very big or very small compared to all other observations), it leaves them out when deciding where to put the whiskers and just plots them directly. Such points are called "outliers," like the book by Malcom Gladwell. We see two such outliers in the plot above.

In case you're interested, R considers an outlier to be any point that is more than 1.5 times the interquartile range away from the box. You don't have to memorize this rule. 

One of the main advantages of boxplots over histograms is that they are simple enough to plot side-by-side. This lets us see how the distribution of a numerical variable is related to a categorical variable. For example, we could see how the distribution of handspan differs for men and women as follows:
```{r}
boxplot(handspan ~ sex, data = survey, ylab= "Handspan (cm)", main = "Handspan by Sex")
```

The syntax used above is different from what we've seen so far but since it will appear again when we look at regression it's worth spending some time to explain. The syntax ``handspan ~ sex`` indicates that we want ``handspan`` *as a function of* ``sex``. The argument ``data = survey`` tells R where to find the variables handspan and sex: they're stored as columns in the dataframe survey.

### ``table``
The command ``table`` is used for making cross-tabs, aka two-way tables, a useful way of summarizing the relationship between two categorical variables. Both arguments of table must be vectors. For example, we can make a cross-tab of eyecolor and sex as follows:
```{r}
table(survey$eye.color, survey$sex)
```
You can also use ``table`` to make a *one-way* crosstab:
```{r}
table(survey$eye.color)
```
Adding the row and column sums is a little clumsy in R: you first need to give your table a name and store it, then use the function ``addmargins``
```{r}
my.table <- table(survey$eye.color, survey$sex)
addmargins(my.table)
```
To convert a cross-tab in R into *percents* rather than counts, we use the function prop.table. This is a little clumsy as well. First, store the original table, then pass it as the argument to ``prop.table`` as you did with ``addmargins`` 
```{r}
my.table <- table(survey$eye.color, survey$sex)
prop.table(my.table)
```
Now you have a table in which percents are expressed as *decimals*. For example, the value of 0.37209 in the row ``Brown`` and the column ``Female`` indicates that about 37% of the people in the dataset are women who have brown eyes. 

To get ordinary percents, we multiply by 100
```{r}
100 * prop.table(my.table)
```
To make this cleaner we can round the result to a desired number of decimal places:
```{r}
round(100 * prop.table(my.table), digits = 1)
round(100 * prop.table(my.table), digits = 0)
```

### Summary Statistics
You've met several of the summary statistics from class in R Tutorial #1. I'll remind you of them here and introduce some others. 

### ``summary``
This command takes a *whole dataframe* as its argument, unlike the other commands for summary statistics. It gives us the mean of each column along with the five-number summary. It also indicates if there are any missing observations (NA's) in the columns:
```{r}
summary(survey)
```

### ``NA``
Remember: R's code for a missing observation is NA. If you try to perform calculations with missing observations, you won't get the answer you expect:
```{r}
sum(survey$height)
```
To tell R to ignore missing observations, set the argument ``na.rm`` to ``TRUE``. This will work with pretty much any of the commands you know for performing mathematical operations on a vector
```{r}
sum(survey$height, na.rm = TRUE)
```

### ``mean``
Calculates the sample mean of a numeric vector
```{r}
mean(survey$height, na.rm = TRUE)
```

### ``median``
Calculates the sample median of a numeric vector
```{r}
median(survey$height, na.rm = TRUE)
```

### ``var``
Calculates the sample variance of a numeric vector
```{r}
var(survey$height, na.rm = TRUE)
```

### ``sd`` 
Calculates the sample standard deviation of a numeric vector
```{r}
sd(survey$height, na.rm = TRUE)
```
This is identical to the (positive) square root of the sample variance:
```{r}
sqrt(var(survey$height, na.rm = TRUE))
```

### ``IQR``
Calculates the interquartile range of a numeric vector (the 75th percentile minus the 25th percentile)
```{r}
IQR(survey$height, na.rm = TRUE)
```

### ``max, min``
These do exactly what they say
```{r}
max(survey$height, na.rm = TRUE)
min(survey$height, na.rm = TRUE)
```
and can be used to calculate the range
```{r}
max(survey$height, na.rm = TRUE) - min(survey$height, na.rm = TRUE)
```
To get both the maximum and minimum at once, you can use the function range;
```{r}
range(survey$height, na.rm = TRUE)
```
Note that this does *not* compute the *summary statistic* called the range. 

### ``quantile``
This function calculates sample quantiles, aka percentiles, of a numeric vector. If you simply pass it a numeric vector with no other arguments, it will give you the five-number summary:
```{r}
quantile(survey$height, na.rm = TRUE)
```
You can ask for specific quantile by using the argument probs
```{r}
quantile(survey$height, na.rm = TRUE, probs = 0.3)
```
Note that quantiles are specified as *probabilities* so the preceding command gives the 30th percentile. You can also ask for multiple quantiles at once:
```{r}
quantile(survey$height, na.rm = TRUE, probs = c(0.1, 0.3, 0.7, 0.9))
```

### ``by``
The function "by" extremely useful: it allows us to calculate summary statistics for a dataframe broken down by the values of a categorical variable (in R parlance, a "factor"). It's also a very flexible function. The best way to get a sense for how it works it to look at some examples. You may also want to take a look at the help file.

First we'll use it to compare the average height of men in the class to that of women:
```{r}
by(survey$height, survey$sex, mean, na.rm = TRUE)
```
Here's an explanation of each of the four arguments:
  1. ``survey$heigh`` is the *numerical variable* that I want to compare across values of a categorical variable. 
  2. ``survey$sex`` is the *categorical variable* (aka a factor) over which I want to compare ``survey$height``
  3. ``mean`` is the summary statistic that I want to compute for ``survey$height``
  4. ``na.rm = TRUE`` is an additional argument that tells the function to ignore missing values.
  
If we wanted to compare the variance of height by sex, we would just change ``mean`` to ``var``  
```{r}
by(survey$height, survey$sex, var, na.rm = TRUE)
```
In fact, you can substitute *any summary statistic* in place of mean. This is why ``by`` is such a powerful function. 

Now, suppose we wanted to compare average height by *eye color*. To do this, simply replace ``survey$sex`` with ``survey$eye.color``
```{r}
by(survey$height, survey$eye.color, mean, na.rm = TRUE)
```

In fact, ``by`` allows us to use *two categorical variables at once* for splitting up a dataframe. Suppose  we wanted to calculate average height by sex *and* credits. In this case, we just change the third argument so that it is a *dataframe* containing the categorical variables we want to use for splitting ``survey$height`` into groups:
```{r}
by(survey$height, survey[,c('sex', 'credits')], mean, na.rm = TRUE)
```
This contains all the information we want, but it's not displayed in a particularly convenient format. There's a neat trick to make this look prettier. It's similar to how we added the column and row sums to the crosstab above. First *assign* the result of your by command to some variable. I'll call mine ``results``
```{r}
results <- by(survey$height, survey[,c('sex', 'credits')], mean, na.rm = TRUE)
```
Then pass this to the function ``as.table``
```{r}
as.table(results)
```
This is much easier to read. The empty cells indicate, for example, that there are no women in the class taking 4.5 credits. They correspond to the NA's in the less pretty version of the same results given above.



### ``cut``
In class we talked about the idea of putting numerical data into "bins" to make a histogram. More generally, it's sometimes necessary to convert numerical data into categorical data. A classic example is converting scores from a course into letter grades. You can do this using the command ``cut``. (In fact, this is the function I'll use to determine your grades at the end of the semester!) First I'll create some fake grade data for us to play with
```{r}
grades <- c(67, 93, 85, 82, 88, 86, 78, 97, 74, 77, 81)
```
The ``cut`` function is a little tricky, so it's worth opening the help file to consult as you read this. Its first argument is a numeric vector: the vector we want to cut. It's second argument gives the "cut points." Let's try this out with some simple grade boundaries:
```{r}
cut(grades, c(60, 70, 80, 90, 100))
```
We can tell that this has created a factor because of the line "Levels: (60,70] ..." that appears in the output. Notice That by default the intervals used in cut are "open on the left, closed on the right." You can reverse this by setting ``right = FALSE``
```{r}
cut(grades, c(60, 70, 80, 90, 100), right = FALSE)
```
This is what we'd use for grade cut-offs since they take the form "X or higher." 

The last step is to add an additional argument to ``cut`` that gives the factor levels meangingful names: in this case, letter grades
```{r}
cut(grades, c(60, 70, 80, 90, 100), c("D", "C", "B", "A"), right = FALSE)
```
Notice that if you used $n$ breakpoints, you need $n-1$ labels since they correspond to the "gaps" between the breaks.


## Exploring the Titanic
Now it's your turn: using what you've learned above, you'll take a closer look at the titanic dataset introduced in R Tutorial #1. You already know enough R to answer some interesting questions, so let's dive in! 


#### A Helpful Hint: Taking the mean of a vector of ones and zeros is the *same thing* as calculating the *proportion of ones*.

1. Read the documentation file for the titanic dataset so that you know what each column in the data represents: http://www.ditraglia.com/econ103/titanic3.txt

2. Download the data, store it in a dataframe called titanic, and display the first few rows, just as you did in R Tutorial #1.
```{r echo=FALSE,results='hide',fig.show='hide'}
#Ex-2
titanic <- read.csv("http://www.ditraglia.com/econ103/titanic3.csv")
head(titanic)
```

3. We'll only be using the following columns of titanic for this example: pclass, survived, sex, age, and fare. Select only these columns and store then in the dataframe titanic, overwriting the original dataset. (Note that I did the same thing above with the survey data.) Display the first few rows to make sure your command worked.
```{r echo=FALSE,results='hide',fig.show='hide'}
#Ex-3
titanic <- titanic[,c('pclass', 'survived', 'sex', 'age', 'fare')]
head(titanic)
```

4. Using the command summary to get an overview of the dataset and answer the following questions:
  - Are there any missing values in this dataset?
  - What was the average age of passengers on the Titanic?
  - What was the average ticket price for passengers on the ship?
  - Do the ticket prices show evidence of skewness? If so, are they positively or negatively skewed?
  - Were there more men or women aboard the Titanic?
  - What proportion of the passengers survived the disaster?
  - To which social class did most of the passengers belong?
```{r echo=FALSE,results='hide',fig.show='hide'}
#Ex-4
summary(titanic)
```
5. Being sure to allow for missing values if necessary, calculate (or plot) and interpret the following:
  - The standard deviation of fares
  - The 90th Percentile of fares 
  - Histogram of fares
```{r echo=FALSE,results='hide',fig.show='hide'}
#Ex-5
sd(titanic$fare, na.rm = TRUE)
quantile(titanic$fare, 0.9, na.rm = TRUE)
hist(titanic$fare)
```
6. Create a boxplot of ``fare`` by ``pclass`` and interpret your results.
```{r echo=FALSE,results='hide',fig.show='hide'}
#Ex-6
boxplot(fare ~ pclass, data = titanic,  main = "Boxplots of fare by pclass")
```
7. Create a dataframe called survivors, containing data for only those passengers who survived the disaster and answer the following questions:
  - What was the average age of the survivors? 
  - Among the survivors, were there more men or women? 
  - Do you notice anything different about social class for survivors compared to all passengers? 
```{r echo=FALSE,results='hide',fig.show='hide'}
#Ex-7
survivors <- subset(titanic, survived == 1)
summary(survivors)
```
8. Who paid more, on average, for passage on the titanic: men or women?
```{r echo=FALSE,results='hide',fig.show='hide'}
#Ex-8
by(titanic$fare,titanic$sex, mean, na.rm = TRUE)
```
9. Create a table of mean fares broken down by ``sex`` and ``pclass``. How does this relate to your answer to the preceding question?
```{r echo=FALSE,results='hide',fig.show='hide'}
#Ex-9
as.table(by(titanic$fare, titanic[,c('sex', 'pclass')], mean, na.rm = TRUE))
```
10. How did the fraction of survivors vary by ``sex`` ``pclass``? Can you think of a possible explanation for the pattern you see in the data?
```{r echo=FALSE,results='hide',fig.show='hide'}
#Ex-10
as.table(by(titanic$survived, titanic[, c("sex", "pclass")], mean))
```
11. Create a categorical variable called ``age.group`` from the numerical variable age using the following cutoffs (http://www.statcan.gc.ca/concepts/definitions/age2-eng.htm). Once you've done this, add ``age.group`` to the dataframe ``titanic`` by using the command ``titanic$age.group <- age.group``
  - Child = [0, 15)  
  - Youth = [15, 25)
  - Adult = [25, 65)
  - Senior = [65, 200)
```{r echo=FALSE,results='hide',fig.show='hide'}
#Ex-11
age.group <- cut(titanic$age, c(0, 15, 25, 65, 200), c("Child", "Youth", "Adult", "Senior"), right = FALSE)
titanic$age.group <- age.group
```
12. Create two tables: one for men and one for women. Each table should show break down survival proportions by pclass and age.group. [Hint: use ``subset`` and ``by``] What do your results suggest? Is there any thing wrong with the way we defined the "Senior" age group in the previous part? 
```{r echo=FALSE,results='hide',fig.show='hide'}
#Ex-12
men <- subset(titanic, sex == 'male')
women <- subset(titanic, sex == 'female')
results.men <- by(men$survived, men[,c("age.group", "pclass")], mean, na.rm = TRUE)
results.women <- by(women$survived, women[,c("age.group", "pclass")], mean, na.rm = TRUE)
round(as.table(results.women),2)
round(as.table(results.men),2)
#It's actually possible to do this without using subset as follows:
full.results <- by(titanic$survived, titanic[c("age.group", "pclass", "sex")], mean, na.rm = TRUE)
round(as.table(full.results),2)
```


