---
title: 'R Tutorial #1 -- Econ 103'
output:
  html_document:
    includes:
      in_header: "google-font.html"
    css: custom.css
    theme: readable
    toc: yes
  pdf_document:
    highlight: zenburn
    toc: yes
---

This is a long tutorial, but the material is fairly straightforward. If you run into any trouble feel free to post on Piazza. 

The most crucial piece of advice for learning a programming language is to recognize it requires the same approach as learning a foreign language -- you'll benefit most from being _actively engaged_ in learning. That means not just reading along with these tutorials, but _actively processing_ what it says and _running the code_ yourself.

Part 1: Installing R
--------------------------------------------------------

Carry out the following two steps *in order*

1. Go to http://cran.r-project.org/ and install the version of R for your operating system.

2. Go to http://rstudio.org/download/desktop and click the link listed under "Recommended for Your System". Follow the instructions to install RStudio.

To make sure this worked, open the program RStudio and go to File > New > R Script. This will open a blank text document. In the document, type the text given in the box below and then click and drag to highlight both lines of code and click the button marked "Run." If everything is working correctly, the console should display ``TRUE``.
```{r initial_test, results='hide'}
x = 5
x == 5
```

Congratulations: you've just written your first R script! To save it, go to File > Save As, and choose a name. NOTE: Always save your scripts as .R files so they'll open in RStudio by default. 

Note that you can run one line of your script at a time by moving your cursor to that line and pressing CONTROL-ENTER or COMMAND-RETURN depending on whether you're running Mac OSX, Linux or Windows. Another helpful shortcut is CONTROL-A  (COMMAND-A on Mac) which highlights all of the lines of code in the text editor.

Part 2: The Absolute Basics
--------------------------------------------------------

Here are some of the most fundamental things you can do with R.

### Arithmetic

```{r arithmetic}
#add numbers
1 + 1
#subtract them
8 - 4
#divide
13/2
#multiply
4*pi
#exponentiate
2^10
```

### Logical Comparison

```{r logicals}
3 < 4
3 > 4
#contrast with 3 = 4; see section about variables below
3 == 4
#!= means "not equal to"
3 != 4
4 >= 5
4 <= 5
2 + 2 == 5
10 - 6 == 4
```

### Strings (text)

Numbers are bread and butter for computers, but text is what will facilitate understanding for us mere mortals.

```{r strings}
'Econometrics is awesome'
#R delimits strings with EITHER double or single quotes.
#  There is only a very minimal difference
"Econometrics is still awesome"
```

### Variables

Just like in algebra, variables are a great form of shorthand. Instead of writing 3.1415926... all the time, we can just write `pi`.

Assignment to a variable happens from right to left -- the value on the right side gets assigned to the _name_ on the left side. You can use nearly anything as a variable name in R. The only rules are:

  1. `.` and `_` are OK, but no other symbols. 
  2. Your variable name must not start with a number or `_` (`2squared` and `_one` are illegal).

[A note for those of you who have programming experience: while R supports object-oriented programming, periods ``.`` do *not* have a special meaning in the language. For historical reasons, R programmers often use periods in place of underscores in variable names, but either works. Just be consistent to keep your code readable.]

```{r variables}
x = 42
x / 2
#if we assign something else to x,
#  the old value is deleted
x = "Melody to Funkytown!"
x
x = 5
x == 5
foo = 3
bar = 5
foo.bar = foo + bar
foo.bar
foo.bar2 = 2 * foo.bar
foo.bar2
foo_bar = foo - bar
foo_bar
```

Note: In programmer speak, `=` here is an "assignment operator" -- it's the thing used to assign values to a variable name. R also has a second assignment operator that you're bound to see sooner or later, `<-`. So `x <- 42` and `x = 42` are _identical_, and both accomplish the task of assigning the value of `42` to the name `x`. We'll try to stick with using `=` since it's easier to type and in some ways more intuitive. See [this](http://stackoverflow.com/questions/1741820/assignment-operators-in-r-and) wonderful post for some more history and a very subtle difference between the two operators that you needn't concern yourself with for now.

### Vectors & Types

In R, a vector is just a (ordered) set of related things. You should basically think of it like a column in Excel.

```{r vectors_1}
x = c(4, 7, 9)
x
y = c('a', 'b', 'c')
y
```

`4`, `7`, and `9` are "related" because they're all numbers; `a`, `b` and `c` are all letters. Having variables is becoming more convenient -- instead of having to write `c(4, 7, 9)` all the time, we can just write `x`.

What happens when we try and combine things that aren't so obviously related?

```{r vectors_2}
x = c(1, TRUE, "three")
x
```

Note the quotation marks. R has converted `1` and `TRUE` into text representations. That's because `1` and `TRUE` are different _type_s than `"three"`. There are four basic types of variables your likely to encounter in this class, listed here in heirarchical order:

  1. `logical`: `TRUE` or `FALSE`
  2. `integer`: `0L`, `-1L`, `1L`, etc. A (real) number without a decimal part. Technical note: they take up less space in the computer than numbers with decimals.
  3. `numeric`: `pi`, `0.34`, `1.4043`, etc. A real number.
  4. `character`: `"some words"`, `"more words"`, etc.

Vectors are converted to the highest number on this list present -- `x` above has `"three"` so the whole vector becomes a `character`.

### Vector Arithmetic and Functions

Vectors make it easy to do many computations all at once -- adding one to a list of numbers, dividing all of them by 3, etc. And as long as two vectors are the same length, we can combine them in natural ways:

```{r vector_arithmetic}
x = c(1, 2, 3)
x + 4
x/3
-x
x^3
y = c(3, 2, 1)
x - y
x * y
x/y
x > 2
x >= 2
```

Just like in math, a function is a way of mapping input to output, and just like in most math classes, you can spot functions since they use parentheses: `()`. We've already seen the _c_oncatenate function `c` used (for example) to create vectors.

We can also apply any number of ubiquitous functions to our vector input.  Just a small taste:

```{r functions}
x = c(1, 2, 3)
#sum: add up the elements of a vector
sum(x)
#Just like you can use the command sum to add up the
#  elements of a numeric vector, you can use
#  prod to take their product:
prod(x)
sqrt(x)
y = c(-1, 2, 4)
#abs: absolute value
abs(y)
#exp: exponential. exp(x) is e^x
exp(y)
#log: _natural_ logarithm (base e)
log(x)
#Note that these functions interpret their input 
#  as *radians* rather than degrees.
sin(x) + cos(y)
max(y)
min(y)
range(y)
mean(x)
median(x)
```

Another thing that we will do all the time is use regularly-spaced sequences of numbers. These are created in R with `:` or `seq`:

```{r sequences}
x = 1:10
x
y = 10:1
y
#some times the gap is not 1
z = seq(0, 1, by = .02)
z
#other times we care less about the gap and more
#  more about how many points we get out
w = seq(0, 1, length.out = 20)
```

In addition to math/arithmetic functions, there is a litany of basic programming functions that you're likely to use all of the time:

```{r programming_functions}
x = 99:32
#length: how many elements (items) are there in x?
length(x)
y = c("hey you!", "out there in the cold")
#what TYPE of variable does R think this is?
class(y)
#rep: repeat/reproduce
rep(y, 4)
#head/tail: display only the beginning/end
#  of an object -- very useful for very
#  large objects
x = 1:100000
head(x)
tail(x)
```

### Subsetting Vectors: `[`

Often we want to examine only part of a vector, most commonly the part of a vector that satisfies some condition, but also looking at the first or last few elements. To do this we _extract_ or subset those elements by using `[`:

```{r vector_extraction_1}
x = c(5, 4, 1)
x[1]
x[3]
x[1:2]
x[2:3]
```

In the syntax `x[something]`, note that _`something` is itself a vector_! So the above is all short-hand for the more complicated types of subsets:

```{r vector_extraction_2}
x = 20:30
x
x[c(1, 3, 5)]
x[c(5, 9)]
x[seq(1, 10, by = 2)]
```

Besides being an `integer`, `something` can be a _logical vector_ of the same length as the vector itself:

```{r vector_extraction_3}
x = c(5, 6, 7)
x[c(TRUE, TRUE, FALSE)]
x[c(FALSE, TRUE, FALSE)]
x[c(FALSE, FALSE, TRUE)]
```

Most commonly we'll do something that's identical to the above but reads more naturally:

```{r vector_extraction_4}
x = c(-1, 0, 1)
x > 0
x[x > 0]
x[x <= 0]
```

We can also _replace_ parts of a vector by subsetting:

```{r vector_replacement}
x = c(-1, 5, 10)
x[3] = 4
x
x[x < 0] = 0
```

### Named Vectors

It's also often useful to _name_ our vectors to help organize the information. Suppose we were keeping track of the ages of the Trumps:

```{r vector_names_1}
trump_ages = c(70, 46, 38, 34, 32, 22, 9)
```

This is nice, but much more useful if we keep track of who each element represents:

```{r vector_names_2}
trump_ages = c(Donald = 70, Melania = 46, Donald_Jr = 38, Ivanka = 34,
               Eric = 32, Tiffany = 22, Barron = 9)
trump_ages
```

We can also use the `names` function to assign names; this is sometimes easier, e.g., if the names have spaces:

```{r vector_names_3}
names(trump_ages) = c("Donald", "Melania", "Donald, Jr.", "Ivanka", "Eric", "Tiffany", "Barron")
trump_ages
```

This also makes code for subsetting much easier to read, since we can subset by the names:

```{r vector_names_4}
trump_ages["Donald"]
trump_ages[c("Donald", "Barron")]
```

### Getting Help: Documentation

If you're unsure of how something works in R -- what the arguments are to a function, how it works, etc. -- your first step is to check the documentation:

```{r documentation}
?sum
?cos
?"="
```

### Lists

We saw above that R doesn't like vectors to have different types: `c(TRUE, 1, "Frank")` becomes `c("TRUE", "1", "Frank")`. But storing objects with different types is absolutely fundamental to data analysis.

R has a different type of object besides a vector used to store data of different types side-by-side: a `list`:

```{r lists_1}
x = list(TRUE, 1, "Frank")
x
```

Note how different the output looks, as compared to using `c`!! The quotation marks are gone except for the last component. You can ignore the mess of `[[` and `[` for now, but as an intimation, consider some more complicated lists:

```{r lists_2}
x = list(c(1, 2), c("a", "b"), c(TRUE, FALSE), c(5L, 6L))
x
y = list(list(1, 2, 3), list(4:5), 6)
y
```

`x` is a list which has 4 components, each of which is a vector with 2 components. This gives the first hint at how R treats a dataset with many variables of different types – at core, R stores a data set in a list!

`y` is a nested list – it’s a list that has lists for some of its components. This is very useful for more advanced operations, but probably won’t come up for quite some time, so don’t worry if you haven’t wrapped your head around this yet.

### Packages

One of the things that makes R truly exceptional is its vast library of user-contributed packages.

R comes pre-loaded with a boat-load of the most common functions / methods of analysis. But in no way is this congenital library complete.

Complementing this core of the most common operations are external _packages_, which are basically sets of functions designed to accomplish specific tasks.

Best of all, unlike some super-expensive programming languages, all of the thousands of packages available to R users (most importantly through CRAN, the **C**omprehensive **R** **A**rchive **N**etwork) are _completely free of charge_. 

The two most important things to know about packages for now is where to find them, how to install them, and how to load them.

We'll work extensively with the `data.table` package, which was built for working with huge data sets.

#### Where to find packages

Long story short: Google. Got a particular statistical technique in mind? The best R package for this is almost always the top Google result if asked correctly.

#### How to install packages

Just use `install.packages`! 

```{r packages_1, eval = FALSE}
install.packages("data.table")
```

This will download the code from the package to your computer to a place that R understands.

We do _not_ yet have access to the functions in the package. We have to load it first.

#### How to load packages

Just add it to your library!

```{r packages_2}
library(data.table)
```

_Et voila_! You'll now have access to all of the awesome functions in the `data.table` package. You can also Google "tutorial data.table" (or in general "tutorial [package name]") and you're very likely to find a trove of sites trying to help you learn the package.

### `data.table`s

Data sets are the lifeblood of a data lover!

As mentioned above, data sets in R basically `list`s where every element has the same length. In basic R, this is done with a `data.frame`, but it'll be easier for a beginner to understand the syntax of a `data.table`, so you can forget about `data.frame`s for now.

We can build a `data.table` from scratch with the `data.table` command. This command lets you build up a `data.table` from several vectors of the same length:

```{r data.table_1}
foo = 1:5
bar = 2 * foo
foo.bar = data.table(foo, bar)
foo.bar
```

In the preceding example I built a `data.table` with only two columns, but you can add as many as you like. Just separate them by commas:

```{r data.table_2}
y = -4:0
data.table(foo, bar, y)
```

### Subsetting data

When you're working with data, you'll often want to look at subsets that satisfy a particular condition. First we'll set up a simple `data.table`:

```{r subsetting_1}
location = c("New York", "Chicago", "Boston", "Boston", "New York")
salary = c(70000, 80000, 60000, 50000, 45000)
title = c("Office Manager", "Research Assistant", "Analyst", "Office Manager", "Analyst")
hours = c(50, 56, 65, 40, 50)
jobsearch = data.table(location, salary, title, hours)
jobsearch
```

Now, suppose you wanted to see only the jobs in New York. You could select them as follows:

```{r subsetting_2}
jobsearch[location == 'New York']
```

Notice the use of the *double equal sign*. This command is *testing a logical condition*. If you use a single equals sign, this won't work since `=` is what is used to name the arguments to a function in R. The preceding command looks at the `data.table` `jobsearch` and then the column `location` and checks which entries satisfy the condition that the `location` is `"New York"`. Finally, the function returns *only these rows* of the `data.table`.

Now suppose you wanted to extract only those jobs that pay more than $50,000. The command for this is as follows:

```{r subsetting_3}
jobsearch[salary > 50000]
```

Finally, suppose the most you're willing to work per week is 50 hours. Here are the jobs you should consider:

```{r subsetting_4}
jobsearch[hours <= 50]
```

### Loading External Data & Data Summary

The vast majority of the time, you won't be using data that you type in by hand -- you'll be importing data from external sources. One of the most common ways to find such data is in comma-separated format -- such files are structured such that each row represents a row of data, and columns are separated by a comma (actually, any separating character is possible), e.g., like this:

```
name,age,company
Mike,24,BCG
Rodrigo,25,Uber
Frank,28,FMC
Ethan,22,AirBnB
```

It's very easy to read files like this into R very quickly using `fread`. The weather site Weather Underground offers lots of historical data in such tabular format. E.g., the data on [this](https://www.wunderground.com/history/airport/KPHL/2017/1/10/MonthlyHistory.html) page about the weather recorded at Philadelphia International Airport is available as a .csv online [here](https://www.wunderground.com/history/airport/KPHL/2017/1/10/MonthlyHistory.html?format=1).

We can read this into R like so:

```{r data_import}
weather = fread('http://michaelchirico.github.io/philly_weather_data.csv')
weather
summary(weather)
names(weather)
```

NB: More typically, instead of a URL, you'll give `fread` the path to where a .csv file is stored on your computer.

### More on `data.table` Access

One of the most fundamental R skills you'll need to learn is how to access parts of a `data.table` or vector. This can be a little confusing at first since there are usually *several different ways to accomplish the same thing*. This section is intended to add more clarity to some of the material on `data.table`s from above. The _best way to learn this is to play around with different commands and see what happens_. There are some exercises at the end of the tutorial to do just that. If you don't get the result you were expecting, try to think about why.

First I'll build a simple `data.table` from the following vectors:

```{r data.table_extra_1}
person = c("Linus", "Snoopy", "Lucy", "Woodstock")
age = c(5, 8, 6, 2)
weight = c(40, 25, 50, 1)
my.data.table = data.table(person, age, weight)
my.data.table
```

#### Fact #1: You can use the same principles to select subsets of vectors and `data.table`s *by position*

The only real difference here is that vectors are one-dimensional
```{r data.table_extra_2}
age[1:2]
age[c(1,3)]
```

whereas `data.table`s are two-dimensional; the first dimension is rows:
```{r data.table_extra_3}
my.data.table[1:2]
my.data.table[c(1, 3)]
```

The second dimension is columns; we can specify rows _and_ columns by giving two numbers inside `[]`:

```{r data.table_extra_4}
#what is the first row of the third column?
my.data.table[1, 3]
#what are the first three rows of the third column?
my.data.table[1:3, 3]
```

If you leave the part *before* the comma blank, you get all the rows:

```{r data.table_extra_5}
my.data.table[ , 2:3]
```

If you leave the part *after* the comma blank (or don't include it at all), you get all the rows:

```{r data.table_extra_6}
my.data.table[c(1,3), ]
my.data.table[c(2,4)]
```

#### Fact #2: There are *three ways* to access the columns of a `data.table` *by name*. 

The first way is to use `[["COLUMN NAME GOES HERE"]]`

```{r data.table_extra_7}
my.data.table[["weight"]]
```

The second is to use `$`, which is often faster to type since it doesn't require the use of quotation marks:

```{r data.table_extra_8}
my.data.table$weight
```

Both of the preceding methods are limited in that they only allow us to reference a single column. We can reference multiple columns as follows:
```{r data.table_extra_9}
my.data.table[ , c("person", "weight")]
```

Since we left the part before the comma blank, this gave us all the rows. We could get the same thing by accessing these columns *by position* (though this is generally not recommended)
```{r data.table_extra_10}
my.data.table[ , 2]
my.data.table[ , c(1,2)]
my.data.table[ , 1:2]
```

In some cases it's easier to access columns of a `data.table` *by name* and in others it's easier to access them *by position*. 

```{r TOGGLE_SOLUTIONS, echo = FALSE, purl = FALSE}
#use this to turn the display of answers on/off
#knitr::opts_chunk$set(echo = FALSE, results = 'hide')
```


Part 3: Exercises
--------------------------------------------------------

**If you can't get R and RStudio to work on your computer, you can do the exercises on the R Fiddle website**

<http://www.r-fiddle.org/#/>

1. Calculate how many minutes there are in a January 

```{r exercise_1}
60 * 24 * 7 * 31
```

2. Add up the numbers 3 1 4 1 5 9 2 6 *without using any plus signs* 
```{r exercise_2}
sum(c(3,1,4,1,5,9,2,6))
```

3. Load the help file for the function `summary`, and use `summary` on an object.
```{r exercise_3}
?summary
summary(iris)
```

4. Suppose I ran the following R commands in order. What result would I get after the fourth command? *Do not* use R to answer this: think it through and then check your answer.
  - `x = 5`
  - `y = 7`
  - `z = x + y`
  - `z + 3 == 15`
  
```{r exercise_4}
x = 5
y = 7
z = x + y 
z + 3 == 15
```

5. How can I get R to print out `"Go Penn!"` thirty times without repeatedly typing this by hand?  
```{r exercise_5}
rep("Go Penn", times = 30)
```

6. Create a vector called `x` containing the sequence -1, -0.9, ... 0, 0.1, ..., 0.9, 1 and then display the result

```{r exercise_6}
x = seq(-1, 1, 0.1)
x
```

7. Create two vectors: `wizards` and `ranking`. The vector `wizards` should contain the following names: Harry, Ron, Fred, George, Sirius. The vector `ranking` should contain the following numbers: 4, 2, 5, 1, 3 in it. Make sure to put these in order.

```{r exercise_7}
#Remember that the elements of character vectors need to be enclosed in quotation marks. Either single or double quotes will work.
wizards = c("Harry", "Ron", "Fred", "George", "Sirius")
ranking = c(4, 2, 5, 1, 3)
```

8. Extract the second element of the vector `wizards`.
```{r exercise_8}
wizards[2]
```

9. Replace the names Fred, George and Sirius in the vector wizards with Hermione, Ginny and Malfoy, respectively. 
```{r exercise_9}
#There are several different ways to do this. Here are two possibilities.
wizards[c(3, 4, 5)] = c("Hermione", "Ginny", "Malfoy")
wizards[3:5] = c("Hermione", "Ginny", "Malfoy")
```

10. Someone who hasn't read Harry Potter needs labels to determine who these characters are. Assign names to the elements of the vector `wizards`: Lead, Friend, Friend, Wife, Rival. Display the result.
```{r exercise_10}
names(wizards) = c("Lead", "Friend", "Friend", "Wife", "Rival")
wizards
```

11. An avid reader of Harry Potter argues that Malfoy is not Harry's rival by the end of the series. Change Rival to Ex-Rival. 
```{r exercise_11}
names(wizards)[5] = "Ex-Rival"
names(wizards)
```

12. In 2009 Steve's income was \$50,000 and his total expenses were \$35,000. In 2010 his income was \$52,000 and his expenses were \$34,000. In 2011, his income was \$52,500 and his expenses were \$38,000. Finally, in 2012 Steve's earnings were \$48,000 and his expenses were \$40,000. Create three vectors to store this information in parallel: `years`, `income` and `expenses`.

```{r exercise_12}
years = c(2009, 2010, 2011, 2012)
income = c(50000, 52000, 52500, 48000)
expenses = c(35000, 34000, 38000, 40000)
```

13. Following on from the previous question, calculate Steve's annual savings and store this in a vector called `savings`. 
```{r exercise_13}
savings = income - expenses
```

14. Assuming zero interest on bank deposits (roughly accurate at the moment), calculate the total amount that Steve has saved over all the years for which we have data.
```{r exercise_14}
sum(savings)
```

15. Create a vector called `z` that lists the numbers from 12 to 23 in *descending order*. 
```{r exercise_15}
z = 23:12
z
```

16. Replace the number 13 with the number 7 in `z`.
```{r exercise_16}
z[z == 13] = 7
z
```

17. Twenty-six students took the midterm. Here are their scores:
18, 95, 76, 90, 84, 83, 80, 79, 63, 76, 55, 78, 90, 81, 88, 89, 92, 73, 83, 72, 85, 66, 77, 82, 99, 87. Assign these values to a vector called `scores`.
```{r exercise_17}
scores = c(18, 95, 76, 90, 84, 83, 80, 79, 63, 76, 55, 78, 90, 81, 88, 89, 92, 73, 83, 72, 85, 66, 77, 82, 99, 87)
```

18. Calculate the mean, median, and range of the scores.
```{r exercise_18}
mean(scores)
median(scores)
range(scores)
```

19. Create three vectors. First store the numeric values 21, 26, 51, 22, 160, 160, 160 in a vector called `age`. Next, store the names Achilles, Hector, Priam, Paris, Apollo, Athena, Aphrodite in a character vector called `person`. Finally store the words Aggressive, Loyal, Regal, Cowardly, Proud, Wise, Conniving in a vector called `description`
```{r exercise_19}
age = c(21, 26, 51, 22, 160, 160, 160)
person = c("Achilles", "Hector", "Priam", "Paris", "Apollo", "Athena", "Aphrodite")
description = c("Aggressive", "Loyal", "Regal", "Cowardly", "Proud", "Wise", "Conniving")
```

20. Create a `data.table` called `trojan.war` whose columns contain the vectors from the previous question.
```{r exercise_20}
trojan.war = data.table(person, age, description)
```

21. Suppose you wanted to display only the column of `trojan.war` that contains each person's `description`. What command would you use? 
```{r exercise_21}
#There are many different ways to do this:
trojan.war[, 3] 
trojan.war$description 
trojan.war[ , "description"]
trojan.war[["description"]]
```

22. What command would you use to show information for Achilles and Hector only?
```{r exercise_22}
#There are several ways to do this. Here are a few:
trojan.war[c(1,2)]
trojan.war[1:2]
#A more advanced way that doesn't require knowing the order of the rows:
trojan.war[person %in% c("Achilles", "Hector")]
```

23. What command would you use to display the `person` and `description` columns for Apollo, Athena and Aphrodite only? 
```{r exercise_23}
#There are many ways to do this. Here are a few:
trojan.war[c(5, 6, 7), c(1, 3)]
trojan.war[5:7, c("person", "description")]
#advanced method
trojan.war[person %in% c("Apollo", "Athena", "Aphrodite"),
           c("person", "description")]
```

24. By now you're probably tired of this data set. A passenger manifest for the Titanic is stored at http://www.ditraglia.com/econ103/titanic3.csv. Read this file and store it in a dataframe called `titanic`.

```{r exercise_24}
titanic = fread("http://www.ditraglia.com/econ103/titanic3.csv")
```

25. Calculate the product of all the even numbers between 2 and 18, inclusive.
```{r exercise_25}
x = seq(2, 18, 2)
x
prod(x)
```

26. The column `survived` in the `titanic` data has a value of "1" to indicate that the passenger in that row survived the disaster. Display only the rows of `titanic` corresponding to passengers that survived.
```{r exercise_26}
titanic[survived == 1]
```

27. The column `sex` in the `titanic` data indicates each passenger's sex. Display only the rows of `titanic` corresponding to men.
```{r exercise_27}
titanic[sex == 'male']
```
