\documentclass[11pt]{article}
\title{Homework and Extra Practice Problems}
\author{Econ 103}

\begin{document}
\maketitle

\section{Introduction to Statistics}
Lectures 1-3\\[0.2in]
Basic definitions, observational and experimental data, taxonomy of variables, summary statistics, samples and populations, z-scores, bias. 
	\begin{itemize}
	\item High priority exercises
	\begin{itemize}
		\item  Problem set II, Part II, exercises 1, 2, 3, 4, 5, 6, 7.
		%\item Midterm 1, Spring 2016, question 2. 
		%\item Midterm 1, Fall 2015, questions 3 and 5. 
		%\item Midterm 1, Spring 2015, questions 1, 2.
		%\item Midterm 1, Fall 2014, question 1 (a, b, c, d), question 4.
		\item Midterm 1, Spring 2014, question 1, 5.
		\item Midterm 1, Fall 2013, question 1, 2, 8, 9.
		\item Midterm 1, Spring 2013, question 1, 4, 6, 7, 9.
		\item Midterm 1, Fall 2012, question 1. 
		%\item Final, Spring 2015, question 2(i)
	\end{itemize} 
	\item Other exercises:
	\begin{itemize}
		\item Book, Chapter 1: 1, 3, 5, 7, 9, 11, 13, 17.
		\item Book, Chapter 2: 1, 3, 5, 7, 13, 15, 17 [bart b, skip the MAD and MSD], 21, 23, 27, 35. 
		\item Midterm 1, Fall 2012, question 2, 3, 6, 11.
		%\item Final, Fall 2015, question 1(a,c)
	\end{itemize}
	\end{itemize}
	



\section{Introduction to Regression}
Lecture 4 \\[0.2in]
Basic notion of regression, how to obtain the coefficient estimates, interpretation of the coefficients, regression to the mean
\begin{itemize}
\item High priority exercises
	\begin{itemize}
		\item Problem set 3, Part II, exercises 1-4.
		%\item Midterm 1, Spring 2016, question 1.
		%\item Midterm 1, Fall 2015, question 2. 
		%\item Midterm 1, Spring 2015, question 3, 6.
		%\item Midterm 1, Fall 2014, question 3.
		\item Midterm 1, Spring 2014, question 2, 6.
		\item Midterm 1, Fall 2013, question 3, 4.
		\item Midterm 1, Spring 2013, question 3, 5.
	\end{itemize}
\item Other exercises
	\begin{itemize}
		\item Book, chapter 11: 1, 3.
		\item Book, chapter 15: 1(a), 5(a).
		\item Midterm 1, Fall 2012, question 4, 5.
		\item Final, Fall 2012, question 1.
	\end{itemize}
\end{itemize}



\section{Basic Probability} 
Lectures 5-7\\[0.2in]
Topics: Definition of probability, terminology, Venn diagrams, axioms of probability, permutations and combinations, independence, multiplication rule, Bayes rule

\begin{itemize}
\item High priority exercises
	\begin{itemize}
		%\item Midterm 1, Spring 2016, questions 3, 4 and 5. 
		%\item Midterm 1, Fall 2015, questions 1, 4, 5, 6. 
		%\item Midterm 1, Spring 2015, questions 4, 5.
		%\item Midterm 1, Fall 2014, question 1 (e, f), question 2,  question 5.
		\item Midterm 1, Spring 2014, question 3, 4.
		\item Midterm 1, Fall 2013, question 5, 6, 7.
		\item Midterm 1, Spring 2013, question 2.
		%\item Midterm 2, Fall 2014, question 1(a, c).
		%\item Final, Fall 2015, question 1(e,f,i).
		%\item Final, Spring 2015, question 1(f, g).
		\item Final, Fall 2013, question 1.
		\item Final, Spring 2013, question 8.
		\item Problem set 4, additional problems, question 1, 5, 6, 8, 9.
	\end{itemize}
\item Other exercises
	\begin{itemize}
		\item Midterm 1, Fall 2012, question 7, 8, 9, 10.
		%\item Final, Fall 2015, question 3(e).
		\item Final, Fall 2012, question 2.
		\item Book, Chapter 3: 1, 3, 5, 9, 11, 13, 15, 17, 19, 21, 23, 25, 27, 29.
	\end{itemize}
\end{itemize}


\section{Discrete Random Variables} 
Lectures 7-10\\[0.2in]
Topics: Definitions (realization, support set), probability mass function, cumulative distribution function, expected value, variance, Bernoulli random variable, Binomial random variable, joint, marginal, conditionals, independence
\begin{itemize}
	\item High priority exercises
	\begin{itemize}
		%\item Midterm 1, Fall 2014, question 1 (g, h, i, j).
		\item Midterm 1, Spring 2013, question 8.
		%\item Midterm 2, Spring 2016, question 1.
		%\item Midterm 2, Fall 2015, question 1(e), 3, 5(a).
		%\item Midterm 2, Spring 2015, question 1(d, h), question 2(b).
		\item Midterm 2, Spring 2014, question 1.
		\item Midterm 2, Fall 2013, question 1.
		\item Midterm 2, Spring 2013, question 1, 2.
		%\item Final, Fall 2015, question 1(j).
		%\item Final, Spring 2015, question 1(h,j).
		%\item Final, Fall 2014, question 2.
		\item Final, Spring 2014, question 1.
		\item Final, Spring 2013, question 3.
		\item Problem set 4, additional problems, question 2, 3, 4, 5, 7, 8, 9.
		\item Problem set 5, additional problems, questions 1, 2, 3, 4, 5, 9, 11, 13, 17.
	\end{itemize}
	\item Other exercises
	\begin{itemize}
		\item Midterm 2, Fall 2012, question 1, 4, 6.
		\item Final, Fall 2012, question 5.
	\end{itemize}
\end{itemize}

\section{Continuous random variables}
Lecture 11-13\\[0.2in]
Topics: Probability mass function, cumulative distribution function, expectations, variance, joint, marginal, conditional, independence. Types of continuous random variables (Uniform, normal, $\chi^2$, student-t, F). 
\begin{itemize}
	\item High priority exercises
		\begin{itemize}
			%\item Midterm 2, Spring 2016, question 2.
			%\item Midterm 2, Fall 2015, question 1(a,b,c), 2, 4, 5, 6.
			%\item Midterm 2, Spring 2015, question 1(c, e, f, g, i, j), question 2, 3, 4.
			%\item Midterm 2, Fall 2014, question 1(b, d, e), question 2(a, c, d), 5, 6.
			\item Midterm 2, Spring 2014, question 2, 3, 6, 7.
			\item Midterm 2, Fall 2013, question 2, 3, 6, 7.
			\item Midterm 2, Spring 2013, question 3, 4, 5, 7.
			%\item Final, Spring 2016, question 1 (up to f), 2.
			%\item Final, Fall 2015, question 2(b,d).
			%\item Final, Spring 2015, question 1(i), 2(a,b,c), 3.
			%\item Final, Fall 2014, question 3.
			\item Final, Fall 2013, question 3.
			\item Final, Spring 2013, question 1, 2.
			\item Problem set 6, additional problems, question 1, 2, 3, 4, 5, 6, 7, 8. 
			\item Problem set 7, additional problems, question 1.
		\end{itemize}
	\item Other exercises
	\begin{itemize}
		\item Midterm 2, Fall 2012, question 2, 3, 7, 8.
		\item Final, Fall 2012, question 3, 4, 7.
		\item Book, Chapter 4, 19, 21, 23.
	\end{itemize}
\end{itemize}

\section{Sampling distribution and estimation}
Lectures 14-15 \\[0.2in]
Topics: Estimator, estimate, sampling distribution, standard error, bias, efficiency, mean-squared error, law of large numbers, central limit theorem.
\begin{itemize}
	\item High priority exercises
		\begin{itemize}
			%\item Midterm 2, Spring 2016, question 3, 4, 5.
			%\item Midterm 2, Spring 2015, question 1(a), 5, 6.
			%\item Midterm 2, Fall 2014, question 1(f), question 4, 7.
			\item Midterm 2, Spring 2014, question 4, 5.
			\item Midterm 2, Fall 2013, question 4, 5.
			\item Midterm 2, Spring 2013, question 6.
			%\item Final, Fall 2015, question 1(g).
			%\item Final, Fall 2014, question 4.
			\item Final, Fall 2013, question 2.
			\item Final, Spring 2013, question 4, 5.
			\item Problem set 7, additional problems, questions 2, 3, 4.
			\item Problem set 9, additional problems, question 1, 3.
			\item Problem set 10, additional problems, question 1.
		\end{itemize}
	\item Other exercises
	\begin{itemize}
		%\item Midterm 2, Fall 2015, question 1(c).
		\item Midterm 2, Fall 2012, question 5, 8 (a-c), 9.
		\item Book, Chapter 6, questions 1, 3, 5, 7, 15, 17, 19(b), 21.
		\item Book, Chapter 7, questions 1, 3, 5, 9, 13, 17, 19.
	\end{itemize}
\end{itemize}

\section{Confidence Intervals}
Lecture 16-19\\[0.2in]
Definitions, confidence level, intuition, confidence intervals for mean and variance, difference in means, confidence interval if population is not normal (central limit theorem), confidence intervals for proportions, independent samples and matched pairs problem, central limit theorem in confidence intervals. 
\begin{itemize}
	\item High priority exercises
		\begin{itemize}
			%\item Midterm 2, Fall 2015, question 1 (f), question 7.
			%\item Midterm 2, Spring 2015, question 1(b).
			%\item Midterm 2, Fall 2014, question 3.
			\item Midterm 2, Spring 2013, question 8.
			%\item Final, Fall 2015, question 1(a), 3(c), 4, 5(a,b,c).
			%\item Final, Spring 2015, question 1(a).
			%\item Final, Fall 2014, question 5(a-d).
			\item Final, Spring 2014, question 3(a,b).
			\item Final, Fall 2013, question 4 (a-c).
			\item Final, Spring 2013, question 6, 9(a-d).
			\item Problem set 8, additional problems, questions 1, 2, 17(c,d),19,21.
			\item Problem set 9, additional questions, question 2, 4, 5, 6, 7.
			\item Problem set 10, additional problems, question 2, 3, 4, 5, 6.
		\end{itemize}
	\item Other exercises
	\begin{itemize}
	 \item Final, Fall 2012, question 6, 8.
	 \item Book, Chapter 8, questions 1, 3, 5, 7, 9.
	\end{itemize}
\end{itemize}


\section{Hypothesis testing}
Lectures 20-23\\[0.2in]
Steps in hypothesis testing, Type I and Type II error, significance level, power of a test, critical value, decision rule, test statistic, one-sided and two-sided alternative, p-value, relationship between hypothesis testing and confidence intervals, hypothesis test using differences, hypothesis testing for proportions.
\begin{itemize}
	\item High priority exercises
		\begin{itemize}
			%\item Final, Spring 2016, question 3, 4.
			%\item Final, Fall 2015, question 1(b,c,d), 5.
			%\item Final, Spring 2015, question 1(b,c,d,e), 4.
			%\item Final, Fall 2014, question 5.
			\item Final, Spring 2014, question 2, 3, 4, 5.
			\item Final, Fall 2013, question 4, 5.
			\item Final, Spring 2013, question 7, 9.
			\item Problem set 11, additional problems, question 1, 2, 3, 4, 5, 6.
		\end{itemize}
	\item Other exercises
	\begin{itemize}
		\item Final, Fall 2012, question 10. 
		\item Book, Chapter 9, questions 23, 27, 29.
	\end{itemize}
\end{itemize}

\section{Regression part II}
Lectures 24--25 \\[0.2in]
Inference in linear regression, confidence intervals, multiple regression, residual standard deviation, R-squared.
testing for proportions.
\begin{itemize}
	\item High priority exercises
		\begin{itemize}
			%\item Final, Spring 2016, question 5.
			%\item Final, Fall 2015, question 6.
			%\item Final, Spring 2015, question 1d, 5.
			%\item Final, Fall 2014, question 1.
			\item Final, Spring 2014, question 6.
			%\item Final, Fall 2014, question 6.
			\item Final, Fall 2013, question 6.
			\item Final, Spring 2013, question 10.
			\item Problem set 12, additional problems, questions 1, 2.
		\end{itemize}
	\item Other exercises
	\begin{itemize}
		\item Book, Chapter 12, questions 2, 3.
		\item Book, Chapter 13, questions 5, 12.
		\item Book, Chapter 13, questions 1, 3, 5. 
	\end{itemize}
\end{itemize}



%\section{Notes}
%\begin{itemize}
%	\item 1.17 Uses confidence intervals very early in the class. I suggest removing it. 
%\end{itemize}


\end{document}


